Semantic integration of physiology phenotypes with an application to the Cellular Phenotype Ontology

### 2012-01-20 trunk/paper/drinking-bioinformatics.tex

\title[Cell Phenotype Ontology]{Semantic integration of physiology
  phenotypes: an application to the cellular phenotype ontology}

@@

\section{Results:}
Here, we present the outline of a theory for an ontology of
physiology-related phenotypes. We provide a formal description of
process attributes and relate them to the attributes of their parts
and participants. We apply our theory to create the Cellular Phenotype
Ontology (CPO). The CPO is an ontology of morphological and
physiological phenotypic abnormalities of cells, cell components and
cellular processes. Its prime application is the unification of
cellular phenotype descriptions across species by providing terms and
uniform definition patterns. The CPO can further be used for the
annotation of observed abnormalities in domains, such as systems
microscopy, in which cellular abnormalities are observed and for which
no phenotype ontology has been created.

@@

In the domain of phenotypes, multiple ontologies have been
developed. In particular, ontologies to characterize mammalian
\cite{Smith2004}, human \cite{Robinson2008}, yeast \cite{ypo} and worm
\cite{wpo} phenotypes are now available, while several more phenotype
ontologies are under development. To benefit from automated reasoning,
integrate phenotypes across species and reuse the content of anatomy
and process ontologies, classes in phenotype ontologies were defined
using the framework of the Phenotypic Attribute and Trait (PATO)
ontology \cite{Gkoutos2005}. According to the PATO framework, a
phenotype can be decomposed, using an Entity-Quality model, into an
affected entity and a quality that characterizes {\em how} the entity
is affected \cite{Gkoutos2005}. Such decompositions have been created
for several widely used phenotype ontologies \cite{Mungall2010,
  Gkoutos2009b, obml2011h1}, and are being applied together with
methods for reusing knowledge contained in anatomy ontologies
\cite{Mungall2010, Hoehndorf2010phene}.

While the PATO framework is now successfully being applied to
semantically integrate phenotypes across species, the diversity and
complexity of phenotypes in which biological processes and functions
are impaired continues to limit the interoperability between phenotype
ontologies. Major challenges for representing and integrating
process-based phenotypes include establishing the link to the
components of biological systems that have the capabilities to exhibit
such a behaviour, and that attributes of processes are often measured
{\em indirectly} and inferred from other attributes.

@@

\begin{methods}
\section{System and methods}
\subsection{Formal ontology}
The ontology as an approach to semantic standardisation was proposed
more than a decade ago and since then has become the dominant
methodology used to semantically categorise phenodeviance.  The
biomedical research community has invested considerable effort and
resources in the development and establishment of ontologies that are
becoming increasingly successful as information management and
integration tools in many disparate scientific fields allowing
interoperability and semantic information processing between diverse
biomedical resources and domains.

In computer science, an ontology is a specification of a
conceptualization of a domain of knowledge \cite{Gruber1995,
  Guarino1998}.  Ontologies commonly distinguish between {\em classes}
(also called {\em concepts}, {\em categories} or {\em universals}) and
{\em individuals} within a domain of knowledge. A class is an entity
that can have {\em instances}, while individuals are entities that
cannot be instantiated \cite{Herre2006}. Examples of individuals
include the Eiffel tower or the 2009 Ironman Triathlon in Hawaii,
while examples of classes include {\em Tower} or {\em Triathlon}. The
Eiffel tower can be an instance of the class {\em Tower}, and the 2009
Ironman Triathlon an instance of {\em Triathlon}.  The meaning of
classes is specified by stating what must be true of their instances.

In addition to classes and individuals, ontologies often include {\em
  relations}. Relations hold between entities, they are the ``the glue
that holds things together, the primary constituents of the facts that
go to make up reality'' \cite{Barwise1989}.

In {\em formal} ontologies, the specification of classes and relations
follows the axiomatic-deductive method. Given a set of terms that are
used within a domain and whose meaning we wish to specify, we begin by
providing {\em explicit definitions} for some terms, potentially
introducing new terms. An explicit definition of a term $t$ is a
statement that can replace every occurrence of $t$ in any sentence.

Eventually, a set of {\em primitive terms} remains that are not
further defined. Following the axiomatic method \cite{Hilbert1918},
using only the primitive terms, we can construct complex
sentences. Based on the intended meaning of the primitive terms, we
consider some of these sentences true and some of them false in our
domain. We select some of the true sentences as {\em axioms} which
provide the core of our ontology. Ideally, the axioms are chosen so
that all true sentences in the domain we intend to represent follow by
means of logical deduction from the axioms. More commonly, however,
only {\em some aspects} of the intended meaning are formally
represented while other aspects are omitted either due to limitations
in language expressivity or due to their irrelevance to the problem
for which an ontology is developed.

Based on the axioms and definitions, we can use deduction to infer
statements that logically follow from the axioms.  The process of
automatically deducing sentences from axioms is called {\em automated
  reasoning}. Automated reasoning allows users of an ontology to carry
out key activities: verifying the ontology's consistency, inferring
hidden knowledge and thereby performing powerful queries.  An ontology
is formally inconsistent if there is a statement $\phi$ such that
$\phi$ and its negation $\neg \phi$ can be inferred from the
ontology's axioms.  If an ontology is formally inconsistent, {\em
  every} statement can be inferred from the ontology.

Automated reasoning can further determine whether classes in an
ontology are unsatisfiable: a class $C$ is unsatisfiable, if it is
impossible for the class to have any instances. Unsatisfiable classes
in an ontology are commonly the result of a contradictory class
definition. % That a class is unsatisfiable is often a special kind of
% unintended consequence that can be drawn from an ontology.

Automated reasoning in the Web Ontology Language (OWL) can be employed
to automatically compute the generalization hierarchy underlying an
ontology as well as for verification of data consistency and complex
queries \cite{Hoehndorf2011incon, Hoehndorf2011models}. Highly
efficient automated reasoners are available to process OWL ontologies
\cite{Sirin2004, Tsarkov2006, Motik2009a}. OWL profiles were developed
to support even large ontologies by further reducing the expressivity
of OWL in order to enable polynomial-time inferences. In particular
the OWL EL profile was found to provide the expressivity required for
most biomedical ontologies \cite{el4, elvira}, and highly optimized
OWL EL reasoners are available or under development to support
reasoning over very large ontologies \cite{el4, cbreasoner}.

A high expressivity is required to accurately specify complex axioms
that constrain the domain under investigation, and languages with
higher expressivity than OWL are often required in the biomedical
domain to achieve this goal \cite{Hoehndorf2009sequences, rnao}. On
the other hand, automated reasoning over large ontologies and
associated datasets benefits from languages with a low complexity of
inferences in which complex axioms cannot be formulated. Therefore, a
possible solution is to use a layered approach: to specify the meaning
of terms using an expressive language, and derive the axioms that
must obtain in a weaker language using deductive inferences.

\subsection{Processes and their participants}
Most biomedical ontologies share common distinctions between different
kinds of entities: physical objects, qualities, functions and
processes. A physical object is an entity that is present as a whole
at a time point, i.e., an entity that has no temporal parts. A quality
is an attribute or feature of an entity. Physical objects together
with their qualities give rise to functions, which are capabilities or
potentials of physical objects. These functions can then be realized
by processes, which are temporally extended entities
\cite{Burek2006}. Examples of classes that may have processes as
instances include {\em Drinking}, {\em Triathlon} or {\em Apoptosis}.
Figure \ref{fig:onto} illustrates these basic categories of being.

Processes commonly have physical objects as participants. Physical
objects are entities that are present as a whole at time points (i.e.,
they have no temporal parts) and may persist through time, i.e., they
may undergo changes during the process \cite{Herre2006, Herre2010}.
We can further distinguish different {\em roles} of participation in a
process \cite{Loebe2007}. For example, a runner may participate in a
{\em Triathlon} process as a {\em Runner (role)}, while another person
can participate in the same process but in a different role (such as
{\em Referee}).

% Amongst the participants of some processes, we will distinguish
% between {\em inputs} and {\em outputs} of processes. Chemical
% reactions, for example, are processes with definite inputs (the {\em
%   reactants}) and outputs ({\em the products}).

\subsection{Gene Ontology}
The Gene Ontology provides a set of ontologies for molecular and
cellular biology, originally designed to support structured
annotations for genes and gene products in all species with respect to
molecular function (MF), biological process (BP) and cellular
component (CC). MF and BP both describe processes, but at different
spatiotemporal scales; in particular, BP includes processes that
unfold within cells and within tissues and organs of multicellular
organisms. Gene product annotations identify participants in the
processes.

@@

PATO allows for the description of affected entities by combining
various ontologies that describe the entities that have been affected,
such as the various anatomical ontologies, GO \cite {Ashburner2000b},
the Cell Type Ontology \cite {Bard2005}, SO \cite {Eilbeck2005} etc
with the various qualities it provides for defining how these entities
were affected.  PATO can be used for annotation either directly in a
so called post-composed (post-coordinated) manner or for providing
formal (logical) definitions (equivalence axioms) to ontologies
containing a set of precomposed (pre-coordinated) phenotype terms. For
instance, to describe the decrease in the length of the sexual cycle
of female animals, we can combine the PATO term \textit{decreased
  duration} ({\tt PATO:0000499}) with the Gene Ontology term
\textit{estrous cycle} ({\tt GO:0042698}), whilst if such a term
existed in a pre-composed ontology (for example the {\tt MP:0009007}
term from the Mammalian Phenotype) it could be used to provide an
equivalence statement between that class and the above PATO-based
description.

\end{methods}

@@

However, not all rates of a process depend on attributes of the
process participants. In particular, a {\em frequency of occurrence}
or {\em event rate} refers to the number of occurrences of a type of
process during a reference process. For example, a {\em rate of heart
  beating} refers to the number of {\em Heart beating} processes that
occur within a reference process (e.g., a process in which the heart
participates with a duration of one minute). Further attributes that
depend on types of processes with regard to a reference process are
{\em distribution patterns}, i.e., how the occurrences of processes of
a particular type are distributed within a reference process. For
example, the heart may beat {\em rhythmically} or {\em arrhythmically}
within a period of time (see Figure \ref{fig:patterns}).

@@

We implement the first two types of abnormality in the CPO. First, as
a subclass of each {\em Abnormality of P} class, we create {\em
  Abnormal duration of P}, which in turn has {\em Increased duration
  of P} and {\em Decreased duration of P} as subclasses. Second, if we
are able to identify {\em inputs} $I(P)$ or {\em outputs} $O(P)$ of
the process $P$ in the formal definitions of the GO, we automatically
generate {\em Abnormality of $I(P)$ in $P$} as well as {\em
  Abnormality of $O(P)$ in $P$}.  The left side of Figure
\ref{fig:overview} illustrates the schema of classes we generate for
single process abnormalities.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

A further domain that will greatly benefit from the CPO is {\em
  systems microscopy}, which aims to understand complex and dynamic
cellular systems by combining automated fluorescence microscopy, cell
microarray platforms, quantitative image analysis and data mining
\cite{Lock2010}.  If we consider some of the studies, which have been
published in this field in the last few years \cite{Neumann2010,
  Schmitz2010, Fuchs2010}, the need for CPO becomes evident.  In the
three studies illustrated in Table \ref{tbl:studies}, live-cell
imaging assays and RNAi knockdown were used to generate phenotypic
profiles that quantify the cellular response to a given siRNA thus
allowing identification of hundreds of genes involved in diverse
biological functions including cell division, migration and survival.
In each study, several phenotypes were detected and described by the
authors without the use of ontologies (see Table \ref{tbl:studies}),
making the integration between datasets extremely difficult. For
example, it is evident that cell division phenotypes were observed in
all three datasets (e.g. {\em Mitotic delay/arrest} and {\em Prolonged
  mitotic exit} or {\em Methaphase delay} and {\em Methaphase cells})
but the overlap between such phenotypes is unclear.

@@

\subsection{Future research}
We implemented the CPO using a pattern-based approach to formulating
phenotypes involving processes. The patterns we identify are based on
pre-existing ontologies, in particular the PATO ontology and the
classification of cellular processes as well as cellular components in
the GO. The result of our method is a large ontology in which classes
for phenotypes are {\em pre-composed}: they are named and defined
within an OWL ontology. However, the large size of the resulting
ontology may impair its utility for data annotation and integration,
and software tools may not always support such very large
ontologies. The alternative to pre-composing all possible phenotype
classes using the patterns we describe is to dynamically generate
appropriately defined classes at the time at which they are being
used. To achieve this goal, software must be developed to support
ontology users in applying these patterns and generate the appropriate
class description when required.

@@

\begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{onto.pdf}
  \caption{Overview over basic top-level categories. Physical objects
    are entities that are wholly present at time points. Qualities are
    attributes of entities. Functions are capabilities that arise from
    physical objects with their qualities, and processes are
    temporally extended entities that may be realizations of a
    function.\label{fig:onto}}
\end{figure}

\begin{figure}
  \centering
  \includegraphics[width=1\textwidth]{processpatterns.pdf}

  \caption{\label{fig:patterns}Six examples of processes with
    non-comparative and comparative process attributes.  We assume
    that the processes labelled $a$, $b$, $c$ and $d$ are all
    instances of the class of processes $P$.  On the left side, three
    regulation (of $P$) processes are illustrated which exhibit
    non-comparative attributes. The first process has an attribute of
    {\em rhythmic} occurrence of $P$ because the instances of $P$ are
    temporally equidistantly distributed. The second example shows an
    {\em arrhythmic} occurrence of $P$, and the third examples shows
    an {\em increasing frequency} (of $P$). A regulation process with
    an increasing frequency (of $P$) is a process in which the
    frequency of occurrences of $P$ is lower in the first half of the
    process than in the second half. The right side of the figure
    illustrates comparative phenotypic descriptions of processes. On
    the upper right, the {\em normal} reference is shown. The second
    example illustrates a {\em late onset} of $P$, i.e., the attribute
    that $P$ processes begin later than {\em normal}. Finally, the
    lower right illustrates a {\em decreased frequency} (of $P$),
    since fewer processes of the type $P$ occur within the reference
    process than {\em normal}.}
\end{figure}

\begin{figure*}
  \centering
  \includegraphics[width=1.2\textwidth, height=.4\textheight, angle=270]{overview.pdf}
  \caption{Overview over the taxonomic structure of CPO. The structure
    is based on a cellular component class $X$ and the cellular
    processes $P(X)$ in which $X$ is involved.\label{fig:overview}}
\end{figure*}

@@

\bibliographystyle{plain}
\bibliography{lc}

>>>>

\title[Physiology phenotypes]{Semantic integration of physiology
  phenotypes with an application to the Cellular Phenotype Ontology}

@@

\section{Results:}
Here, we present the outline of a theory and its implementation for an
ontology of physiology-related phenotypes. We provide a formal
description of process attributes and relate them to the attributes of
their temporal parts and participants. We apply our theory to create
the Cellular Phenotype Ontology (CPO). The CPO is an ontology of
morphological and physiological phenotypic characteristics of cells,
cell components and cellular processes. Its prime application is the
unification of cellular phenotype descriptions across species by
providing terms and uniform definition patterns. The CPO can further
be used for the annotation of observed abnormalities in domains, such
as systems microscopy, in which cellular abnormalities are observed
and for which no phenotype ontology has been created.

@@

In the domain of phenotypes, multiple ontologies have been
developed. In particular, ontologies to characterize mammalian
\cite{Smith2004}, human \cite{Robinson2008}, yeast \cite{ypo} and worm
\cite{wpo} phenotypes are now available, while several more phenotype
ontologies are under development. To benefit from automated reasoning,
integrate phenotypes across species and reuse the content of anatomy
and process ontologies, we have defined process classes using the
framework of the Phenotypic Attribute and Trait (PATO) ontology
\cite{Gkoutos2005}. According to the PATO framework, a phenotype can
be decomposed, using an Entity-Quality model, into an affected entity
and a quality that characterizes {\em how} the entity is affected
\cite{Gkoutos2005}. Such decompositions have been created for several
widely used phenotype ontologies \cite{Mungall2010, Gkoutos2009b,
  obml2011h1}, and are being applied together with methods for reusing
knowledge contained in anatomy ontologies \cite{Mungall2010,
  Hoehndorf2010phene}.

While the PATO framework is now successfully being applied to
semantically integrate phenotypes across species, the diversity and
complexity of phenotypes in which biological processes and functions
are impaired continues to limit the interoperability between phenotype
ontologies. Major challenges for representing and integrating process
phenotypes include establishing the link to the components of
biological systems that have the capabilities to exhibit such a
behaviour, and that attributes of processes are often measured {\em
  indirectly} and inferred from other attributes.

@@

\begin{methods}
\section{System and methods}
% \subsection{Formal ontology}
% The ontology as an approach to semantic standardisation was proposed
% more than a decade ago and since then has become the dominant
% methodology used to semantically categorise phenodeviance.  The
% biomedical research community has invested considerable effort and
% resources in the development and establishment of ontologies that are
% becoming increasingly successful as information management and
% integration tools in many disparate scientific fields allowing
% interoperability and semantic information processing between diverse
% biomedical resources and domains.

% In computer science, an ontology is a specification of a
% conceptualization of a domain of knowledge \cite{Gruber1995,
%   Guarino1998}.  Ontologies commonly distinguish between {\em classes}
% (also called {\em concepts}, {\em categories} or {\em universals}) and
% {\em individuals} within a domain of knowledge. A class is an entity
% that can have {\em instances}, while individuals are entities that
% cannot be instantiated \cite{Herre2006}. Examples of individuals
% include the Eiffel tower or the 2009 Ironman Triathlon in Hawaii,
% while examples of classes include {\em Tower} or {\em Triathlon}. The
% Eiffel tower can be an instance of the class {\em Tower}, and the 2009
% Ironman Triathlon an instance of {\em Triathlon}.  The meaning of
% classes is specified by stating what must be true of their instances.

% In addition to classes and individuals, ontologies often include {\em
%   relations}. Relations hold between entities, they are the ``the glue
% that holds things together, the primary constituents of the facts that
% go to make up reality'' \cite{Barwise1989}.

% In {\em formal} ontologies, the specification of classes and relations
% follows the axiomatic-deductive method. Given a set of terms that are
% used within a domain and whose meaning we wish to specify, we begin by
% providing {\em explicit definitions} for some terms, potentially
% introducing new terms. An explicit definition of a term $t$ is a
% statement that can replace every occurrence of $t$ in any sentence.

% Eventually, a set of {\em primitive terms} remains that are not
% further defined. Following the axiomatic method \cite{Hilbert1918},
% using only the primitive terms, we can construct complex
% sentences. Based on the intended meaning of the primitive terms, we
% consider some of these sentences true and some of them false in our
% domain. We select some of the true sentences as {\em axioms} which
% provide the core of our ontology. Ideally, the axioms are chosen so
% that all true sentences in the domain we intend to represent follow by
% means of logical deduction from the axioms. More commonly, however,
% only {\em some aspects} of the intended meaning are formally
% represented while other aspects are omitted either due to limitations
% in language expressivity or due to their irrelevance to the problem
% for which an ontology is developed.

% Based on the axioms and definitions, we can use deduction to infer
% statements that logically follow from the axioms.  The process of
% automatically deducing sentences from axioms is called {\em automated
%   reasoning}. Automated reasoning allows users of an ontology to carry
% out key activities: verifying the ontology's consistency, inferring
% hidden knowledge and thereby performing powerful queries.  An ontology
% is formally inconsistent if there is a statement $\phi$ such that
% $\phi$ and its negation $\neg \phi$ can be inferred from the
% ontology's axioms.  If an ontology is formally inconsistent, {\em
%   every} statement can be inferred from the ontology.

% Automated reasoning can further determine whether classes in an
% ontology are unsatisfiable: a class $C$ is unsatisfiable, if it is
% impossible for the class to have any instances. Unsatisfiable classes
% in an ontology are commonly the result of a contradictory class
% definition. % That a class is unsatisfiable is often a special kind of
% % unintended consequence that can be drawn from an ontology.

% Automated reasoning in the Web Ontology Language (OWL) can be employed
% to automatically compute the generalization hierarchy underlying an
% ontology as well as for verification of data consistency and complex
% queries \cite{Hoehndorf2011incon, Hoehndorf2011models}. Highly
% efficient automated reasoners are available to process OWL ontologies
% \cite{Sirin2004, Tsarkov2006, Motik2009a}. OWL profiles were developed
% to support even large ontologies by further reducing the expressivity
% of OWL in order to enable polynomial-time inferences. In particular
% the OWL EL profile was found to provide the expressivity required for
% most biomedical ontologies \cite{el4, elvira}, and highly optimized
% OWL EL reasoners are available or under development to support
% reasoning over very large ontologies \cite{el4, cbreasoner}.

% A high expressivity is required to accurately specify complex axioms
% that constrain the domain under investigation, and languages with
% higher expressivity than OWL are often required in the biomedical
% domain to achieve this goal \cite{Hoehndorf2009sequences, rnao}. On
% the other hand, automated reasoning over large ontologies and
% associated datasets benefits from languages with a low complexity of
% inferences in which complex axioms cannot be formulated. Therefore, a
% possible solution is to use a layered approach: to specify the meaning
% of terms using an expressive language, and derive the axioms that
% must obtain in a weaker language using deductive inferences.

% \subsection{Processes and their participants}
% Most biomedical ontologies share common distinctions between different
% kinds of entities: physical objects, qualities, functions and
% processes. A physical object is an entity that is present as a whole
% at a time point, i.e., an entity that has no temporal parts. A quality
% is an attribute or feature of an entity. Physical objects together
% with their qualities give rise to functions, which are capabilities or
% potentials of physical objects. These functions can then be realized
% by processes, which are temporally extended entities
% \cite{Burek2006}. Examples of classes that may have processes as
% instances include {\em Drinking}, {\em Triathlon} or {\em Apoptosis}.
% Figure \ref{fig:onto} illustrates these basic categories of being.

% Processes commonly have physical objects as participants. Physical
% objects are entities that are present as a whole at time points (i.e.,
% they have no temporal parts) and may persist through time, i.e., they
% may undergo changes during the process \cite{Herre2006, Herre2010}.
% We can further distinguish different {\em roles} of participation in a
% process \cite{Loebe2007}. For example, a runner may participate in a
% {\em Triathlon} process as a {\em Runner (role)}, while another person
% can participate in the same process but in a different role (such as
% {\em Referee}).

% % Amongst the participants of some processes, we will distinguish
% % between {\em inputs} and {\em outputs} of processes. Chemical
% % reactions, for example, are processes with definite inputs (the {\em
% %   reactants}) and outputs ({\em the products}).

\subsection{Gene Ontology}
The Gene Ontology provides a set of ontologies for molecular and
cellular biology, originally designed to support structured
annotations for genes and gene products in all species with respect to
molecular function (MF), biological process (BP) and cellular
component (CC). MF and BP both describe processes, but at different
spatiotemporal scales; in particular, BP includes processes that
unfold within cells and within tissues and organs of multicellular
organisms. Gene product annotations can be interpreted as identifying
participants in the processes.

@@

PATO allows for the description of affected entities by combining
various ontologies that describe the entities that have been affected,
such as the various anatomical ontologies, GO \cite {Ashburner2000b},
the Cell Type Ontology \cite {Bard2005}, SO \cite {Eilbeck2005} etc
with the various qualities it provides for defining how these entities
were affected.  PATO can be used for annotation either directly in a
so called post-composed (post-coordinated) manner or for providing
formal (logical) definitions (equivalence axioms) to ontologies
containing a set of precomposed (pre-coordinated) phenotype terms. For
instance, to describe the decrease in the length of the sexual cycle
of female animals, we can combine the PATO term \textit{decreased
  duration} ({\tt PATO:0000499}) with the Gene Ontology term
\textit{estrous cycle} ({\tt GO:0042698}), whilst if such a term
existed in a pre-composed ontology (for example the {\tt MP:0009007}
term from the Mammalian Phenotype) it could be used to provide an
equivalence statement between that class and the above PATO-based
description.
\end{methods}

@@

However, not all rates of a process depend on attributes of the
process participants. In particular, a {\em frequency of occurrence}
or {\em event rate} refers to the number of occurrences of a type of
process during a reference process. For example, a {\em rate of heart
  beating} refers to the number of {\em Heart beating} processes that
occur within a reference process (e.g., a process in which the heart
participates with a duration of one minute). Further attributes that
depend on types of processes with regard to a reference process are
{\em distribution patterns}, i.e., how the occurrences of processes of
a particular type are distributed within a reference process. For
example, the heart may beat {\em rhythmically} or {\em arrhythmically}
within a period of time (see Figure \ref{fig:patterns}).
\begin{figure}
  \centering
  \includegraphics[width=.5\textwidth]{processpatterns.pdf}
  \caption{\label{fig:patterns}Six examples of processes with
    non-comparative and comparative process attributes.  We assume
    that the processes labelled $a$, $b$, $c$ and $d$ are all
    instances of the class of processes $P$.  On the left side, three
    regulation (of $P$) processes are illustrated which exhibit
    non-comparative attributes. The first process has an attribute of
    {\em rhythmic} occurrence of $P$ because the instances of $P$ are
    temporally equidistantly distributed. The second example shows an
    {\em arrhythmic} occurrence of $P$, and the third examples shows
    an {\em increasing frequency} (of $P$). A regulation process with
    an increasing frequency (of $P$) is a process in which the
    frequency of occurrences of $P$ is lower in the first half of the
    process than in the second half. The right side of the figure
    illustrates comparative phenotypic descriptions of processes. On
    the upper right, the {\em normal} reference is shown. The second
    example illustrates a {\em late onset} of $P$, i.e., the attribute
    that $P$ processes begin later than {\em normal}. Finally, the
    lower right illustrates a {\em decreased frequency} (of $P$),
    since fewer processes of the type $P$ occur within the reference
    process than {\em normal}.}
\end{figure}

@@

We implement the first two types of abnormality in the CPO. First, as
a subclass of each {\em Abnormality of P} class, we create {\em
  Abnormal duration of P}, which in turn has {\em Increased duration
  of P} and {\em Decreased duration of P} as subclasses. Second, if we
are able to identify {\em inputs} $I(P)$ or {\em outputs} $O(P)$ of
the process $P$ in the formal definitions of the GO, we automatically
generate {\em Abnormality of $I(P)$ in $P$} as well as {\em
  Abnormality of $O(P)$ in $P$}.  The left side of Figure
\ref{fig:overview} illustrates the schema of classes we generate for
single process abnormalities.
\begin{figure*}
  \centering
%  \includegraphics[width=1.2\textwidth, height=.4\textheight, angle=270]{overview.pdf}
  \includegraphics[width=1.1\textwidth]{overview.pdf}
  \caption{Overview over the taxonomic structure of CPO. The structure
    is based on a cellular component class $X$ and the cellular
    processes $P(X)$ in which $X$ is involved.\label{fig:overview}}
\end{figure*}

@@

A further domain that will greatly benefit from the CPO is {\em
  systems microscopy}, which aims to understand complex and dynamic
cellular systems by combining automated fluorescence microscopy, cell
microarray platforms, quantitative image analysis and data mining
\cite{Lock2010}.  If we consider some of the studies, which have been
published in this field in the last few years \cite{Neumann2010,
  Schmitz2010, Fuchs2010}, the need for CPO becomes evident.  In the
three studies, live-cell imaging assays and RNAi knockdown were used
to generate phenotypic profiles that quantify the cellular response to
a given siRNA thus allowing identification of hundreds of genes
involved in diverse biological functions including cell division,
migration and survival.  In each study, several phenotypes were
detected and described by the authors without the use of ontologies or
controlled vocabularies, making the integration between datasets
extremely difficult. For example, it is evident that cell division
phenotypes were observed in all three datasets and referred to by
terms such as {\em Mitotic delay/arrest}, {\em Prolonged mitotic
  exit}, {\em Methaphase delay} and {\em Methaphase cells}). Without a
controlled vocabulary of cellular phenotypes, the overlap between such
phenotype descriptions is unclear.

@@

\subsection{Future research}
Our main contribution is an analysis of process phenotypes that are
used across multiple domains and scales and which are crucial for
understanding and representing physiology of living systems. The
Ontology of Physics in Biology (OPB) \cite{opb} is an ontology that
has recently been proposed to characterize physiological processes and
the physical qualities of biological entities based on a theory of
fluid dynamics. It is an important goal for future research to
incorporate the OPB in phenotypic descriptions and make our theory of
process phenotypes compatible with the physical descriptions of
processes and their attributes as outlined by the OPB.

We implemented the CPO using a pattern-based approach to formulating
phenotypes involving processes. The patterns we identify are based on
pre-existing ontologies, in particular the PATO ontology and the
classification of cellular processes as well as cellular components in
the GO. The result of our method is a large ontology in which classes
for phenotypes are {\em pre-composed}: they are named and defined
within an OWL ontology. However, the large size of the resulting
ontology may impair its utility for data annotation and integration,
and software tools may not always support such very large
ontologies. The alternative to pre-composing all possible phenotype
classes using the patterns we describe is to dynamically generate
appropriately defined classes at the time at which they are being
used. To achieve this goal, software must be developed to support
ontology users in applying these patterns and generate the appropriate
class description when required.

@@

@@

\bibliographystyle{plain}
\bibliography{/home/leechuck/Documents/papers/bibtex/lc}
